% $Header$
% MBDyn (C) is a multibody analysis code.
% http://www.mbdyn.org
%
% Copyright (C) 1996-2008
%
% Pierangelo Masarati  <masarati@aero.polimi.it>
%
% Dipartimento di Ingegneria Aerospaziale - Politecnico di Milano
% via La Masa, 34 - 20156 Milano, Italy
% http://www.aero.polimi.it
%
% Changing this copyright notice is forbidden.
%
% This program is free software; you can redistribute it and/or modify
% it under the terms of the GNU General Public License as published by
% the Free Software Foundation (version 2 of the License).
% 
%
% This program is distributed in the hope that it will be useful,
% but WITHOUT ANY WARRANTY; without even the implied warranty of
% MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
% GNU General Public License for more details.
%
% You should have received a copy of the GNU General Public License
% along with this program; if not, write to the Free Software
% Foundation, Inc., 59 Temple Place, Suite 330, Boston, MA  02111-1307  USA

\documentclass[10pt,dvips]{report}

%\usepackage[pdftex]{graphicx}
\usepackage[T1]{fontenc}
\usepackage{ae,aecompl}
\usepackage{graphicx}
\usepackage{psfrag}
\usepackage{amsmath}
\usepackage[dvips,breaklinks=true,colorlinks=false]{hyperref}
\usepackage{html}
\usepackage{comment}

\input{stdmacro}

\newcommand{\kw}[1]{\texttt{#1}}
%\newcommand{\kw}[1]{\fbox{\texttt{#1}}}
\newcommand{\T}[1]{\boldsymbol{#1}}
\newcommand{\TT}[1]{\boldsymbol{#1}}

% Custom
\topmargin 0.31cm % 0.00cm
\oddsidemargin 0.00cm
\evensidemargin 0.00cm
\marginparsep 0.00cm
\textwidth 15.92cm
\textheight 23.00cm % 23.62cm

\begin{document}

\begin{latexonly}
\title{\bf MBDyn Input File Format \\
Version
\input{../build/version}
}
\author{Pierangelo Masarati \vspace{5mm}\\
    \sc Dipartimento di Ingegneria Aerospaziale \\
    \sc Politecnico di Milano}
\date{Automatically generated \today}
\maketitle
\end{latexonly}

\begin{htmlonly}
\begin{center}
\textbf{\LARGE MBDyn Input File Format}

\emph{\large Pierangelo Masarati}

\textsc{Dipartimento di Ingegneria Aerospaziale \\ Politecnico di Milano}

\today
\end{center}
\end{htmlonly}

\tableofcontents
\newpage
\listoffigures
\newpage
\listoftables
\newpage

\chapter*{Introduction}
This document describes the format of the input data for MBDyn,
the \underline{M}ulti\underline{b}ody \underline{Dyn}amics analysis suite.
It should be considered a syntax more than a format, since the rules of the
input allow a lot of freedom in the file format. 

As the title states, this manual describes the syntax of the input.
The reader should not expect to \emph{learn} how to model systems
from this text; it is rather a reference manual where the user
can find the exact and detailed description of the syntax of some
entity, where it is assumed that the user already knows that entity
exactly does what is required to model that system or solve that problem.

To get on the right track, one may have a look at the 
\emph{Frequently Asked Questions} (Appendix~\ref{sec:FAQ}),
a steadily growing chapter that collects the most common issues
arising from users.
They might help pointing to the right statement for a given purpose,
but in any case nothing can replace the judgment of the user
about what entity exactly fits some need.
The only document that may help users in growing this type of judgment
is the \emph{tutorials} book, available from the website.
For more specific questions, one should try the \kw{mbdyn-users@mbdyn.org}
mailing list (browse the archives, just in case, then post a question;
posting is restricted to subscribers; subscription is free).

\section*{Conventions on the Notation}
Throughout the manual, a (sometimes na\"{\i}ve) 
Backus-Naur-like syntax description is used. 
Extensions are made, such as to put optional arguments in square brackets
\kw{[]}, mutually exclusive arguments in curly brackets \kw{\{\}},
separated by the logical ``or'' \kw{|} operator.
Non-terminal symbols are enclosed in angle brackets \kw{<>}, while
terminal symbols are written in normal types.
The association of a non-terminal with terminal or non-terminal
symbols is performed by means of the operator `\kw{::=}'. 
When required, the type of a non-terminal symbol is enforced in a ``C''
programming language casting style, by prepending a (usually
self-explanatory) type enclosed in plain brackets \kw{()}.


\section*{Remarks}
The input of the program MBDyn is subject to continuous changes
at a fast pace, since the code is under development.
This documentation attempts to give some insight into the logics 
that drove its implementation.
While most of the software resulted from a careful design, 
some portions are ``as they are'' only because they were made in a hurry, 
or because no better way was at hand at the moment they were made.
The input format partially reflects the development of the software.
Whenever changes in the input format and in the syntax 
of existing parts are required, an effort will be attempted to make 
it as much backward compatible as possible, or at least reliable 
diagnostics and error checking will be put in place that warns 
for possible erroneous inputs related to a change in the input format. 
At present the input of MBDyn already warns for many possible errors 
and makes as many cross-checks as possible on the consistency of the data. 
Nearly every warning and error issue shows the line of the input file 
where the error occurred. 
This document will be subjected to changes, so be sure you always have 
a release that is aligned with the version of the code you're running.

For any question or problem, to fix typos, bugs, for comments and
suggestions, please contact the Development Team
without hesitation:\vspace{10mm}\\

\noindent
\begin{tabular}{ll}
\multicolumn{2}{l}{Pierangelo Masarati,} \\
\multicolumn{2}{l}{MBDyn Development Team} \\
\multicolumn{2}{l}{Dipartimento di Ingegneria Aerospaziale} \\
\multicolumn{2}{l}{Politecnico di Milano} \\
\multicolumn{2}{l}{via La Masa 34, 20156 Milano, Italy} \\
Fax: & +39 02 2399 8334 \\
E-mail: & \htmladdnormallink{\kw{mbdyn@aero.polimi.it}}{mailto:mbdyn@aero.polimi.it} \\
Web: & \htmladdnormallink{\kw{http://www.aero.polimi.it/\~{}mbdyn/}}{http://www.aero.polimi.it/~mbdyn/}
\end{tabular}
\vspace{10mm}


\noindent
This manual is also available online at \\
\htmladdnormallink{\kw{http://www.aero.polimi.it/\~{}masarati/MBDyn-input/manual/index.html}}{http://www.aero.polimi.it/~masarati/MBDyn-input/manual/index.html}

\noindent
Make sure you use the manual that refers to the version of MBDyn 
that you are using; from 1.2 on, the input manual is marked
with the same version number of the package.

\noindent
The website \htmladdnormallink{\kw{http://www.aero.polimi.it/\~{}mbdyn/}}{http://www.aero.polimi.it/~mbdyn/}
may advertise different versions of the manual, including those related 
to past releases and a HEAD or -devel version, which refers to current
development; this can be useful to check whether the current
development may address problems you are having and, since
the updating of the manual is not as fast as code releases,
it may describe features that are already in the most recent 
release.

\input{overview}
\input{general}
\input{data}
\input{integrators}
\input{controldata}
\input{nodes}
\input{drivers}
\input{elements}

\appendix
\input{netcdfout}
\input{modal-fem-format}
\input{outputresults} 
\input{logfile} 
\input{faq}

\bibliographystyle{unsrt}
\bibliography{mybib,input}


\end{document}
